\chapter{Foundations}
\label{ch:foundations}

The phenomena to be analyzed in this thesis include all methods to structure
and to describe digital data. To experience these phenomena we must first
broaden our view to see where they can be found. For this reason, this chapter
will first introduce the disciplines that deal with data and the description of
digital documents. This introduction also includes definitions of some basic
concepts and notations. These foundations are both used during collection and
analysis in the proceeding chapters and they can be instances of phenomena in
their own right. For instance mathematical set theory is used to define other
methods of data structuring, but set theory alone can also be used for data
structuring. We cannot fully avoid this circularity as every description must
be formulated in some other description --- basically this is the core problem
of data description. Nevertheless we can show how different disciplines
approach and tackle the problem. The largest part of this chapter introduces 
mathematics (section~\ref{sec:mathematics}) and computer science
(section~\ref{sec:informatics}) because these make the traditional foundations
of data: mathematics has proved to be an effective tool to exactly describe
structures of any kind and computer science provides the most examples, as most
problems of practical data processing belong to its domain. The approach of
library and information science (section~\ref{sec:lis}) is different: it is
basically concerned with the organization and description of documents. While
more and more documents become digital, the discipline should more and more
deal with data. The impact of philosophy is more subtle: as outlined in
section~\ref{sec:philosophy} there is not much explicit philosophy of data, but
philosophical issues permeate all other disciplines and philosophy helps to
reveal blind spots of other points of views. Semiotics
(section~\ref{sec:semiotics}) is relevant to this thesis because it deals with
signs and language, which all meaningful data is an instance of.
Section~\ref{sec:patterntheory} finally introduces the fundamental concepts
of patterns and pattern languages.  Both can be combined to pattern theory,
which is more a practice or an art than a scientific discipline.

\include{sec-mathematics}

\include{sec-computerscience}

\include{sec-lis}

\include{sec-philosophy}

\include{sec-semiotics}

\include{sec-patterndesign}


