%% Title as required by Humboldt University

% enable roman page numbers
\renewcommand{\thepage}{\roman{page}}

\makeatletter

% title page
\begin{titlepage}

\begin{center}
\vfill

{\bfseries
  {\huge\strut\ignorespaces\@title\par}
  {\large\strut\ignorespaces\@subtitle\par}
}\vfill

%\begin{otherlanguage}{ngerman}
{\huge\textsc{Dissertation}}\par
{\large in support of the degree of\par
\vskip 1.5\baselineskip
{\it Doctor philosophiae\\(Dr. phil.)} \par
\vskip 1.5\baselineskip
by \@author
\vskip 1.5\baselineskip
submitted at January 7\textsuperscript{th} 2013\par
defended at May 31\textsuperscript{st} 2013\par
\vskip 1\baselineskip
at the Faculty of Philosophy I\\
Humboldt-University Berlin\\
Berlin School of Library and Information Science\par
\vskip 1\baselineskip
Reviewer:\\
Prof. Dr. Stefan Gradman \\
Prof. Dr. Felix Sasaki \\
Prof. William Honig, PhD
}
%\end{otherlanguage}
\end{center}

\pagebreak
\thispagestyle{empty}
\vfill

\noindent This document is licensed under the terms of the Creative Commons
Attribution-ShareAlike license (CC-BY-SA). Feel free to reuse any parts of it
as long as attribution is given to Jakob Vo\ss{} and the result is licensed under
CC-BY-SA as well.

\vskip 1\baselineskip

\noindent
The full source code of this document, its variants and corrections are
available at \url{https://github.com/jakobib/phdthesis2013}.  Selected parts
and additional content are made available at \url{http://aboutdata.org}

\vskip 1\baselineskip

\noindent
A digital copy of this thesis (with same pagination but larger margins to fit
A4 paper format) is archived at \url{http://edoc.hu-berlin.de/}.

\vskip 1\baselineskip

\noindent
A printed version is published through CreateSpace and available by Amazon and
selected distributors.

\vskip 1\baselineskip

\noindent
ISBN-13: 978-1-4909-3186-9\\
ISBN-10: 1-4909-3186-4

\vskip 1\baselineskip\noindent
Cover: the Arecibo message, sent into empty space in 1974 (image
CC-BY-SA Arne Nordmann, \url{http://commons.wikimedia.org/wiki/File:Arecibo_message.svg})

\end{titlepage}

\makeatother
